Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method

### 2023-05-02 manu-draft.qmd
rework text about apa actual comparisons, no more slope/intercept analyses

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were observations from the year 2012 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $r$, and $b$.  All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].  Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

@@

The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D). Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and linear regression fits to evaluate potential biases in each method. For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates). Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each. Root mean square errors (RMSE) of each linear model were also evaluated. Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of method or the input time series, were positive and significantly correlated, as expected. The intercept values were all significant, excluding P with the observed dissolved oxygen, suggesting some level of bias when comparing metabolic estimates between the methods. The slope estimates for R and D were significantly different for metabolic estimates from the observed dissolved oxygen, where all but R were significantly different using the detided dissolved oxygen. The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed. This result suggests that detiding the time series produces more comparable metabolic estimates between methods.

As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) than using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for P and positive values for R, i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates was much larger with observed dissolved oxygen. Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.

@@

The EBASE R package can also accommodate missing observations in the input data. This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.).  The R package accommodates data gaps using linear interpolation for any required parameter used by EBASE.  This allows the Bayesian routine to provide continuous estimates across the time series.  The results are then post-processed such that any metabolic estimates that exceed a threshold for interpolated observations are removed from the output.  The analyst can determine the appropriate length of time to use for excluding results based on interpolated values.  The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours.  Although this does not apply to the synthetic time series, some of the estimates using the 2012 observed data for Apalachicola Bay were excluded from the analysis.  This feature allows EBASE to be used more readily because an analyst does not need to manually screen data that include missing values.

@@

```{r}
#| label: tbl-apacmptab
#| tbl-cap: "Summary statistics of daily metabolic estimates comparing the EBASE and Odum methods for net ecosystem metabolism (NEM), gross primary production (P), respiration (R), and gas exchange (D). Separate comparisons are made for estimates based on observed and detided dissolved oxygen. Summary statistics include Pearson correlation coefficients and intercept, slope, and root mean square error (RMSE) of a linear fit between the same metabolic estimates obtained from each method. Significance of the p-values for the intercept and slope indicates if the estimate is different from zero or one, respectively."
apacmptab
```

>>>>

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $r$, and $b$.  All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].  Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

@@

The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D). Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and root mean square errors (RMSE) of the linear regression fit comparing the same metabolic parameter between methods. Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected.  However, correlations were generally stronger for the metabolic results based on the detided dissolved oxygen time series.  The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results. @fit-apacmpfig shows that the Odum results for gas exchange were more variable for the same values returned as no or small gas exchange for the EBASE method based on the detided dissolved oxygen. The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.

The comparisons in @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` suggest that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified by stronger correlations and lower RMSE using the detided dissolved oxygen.  Further, effects of tidal advection can be seen with estimates from the Odum method as negative values for P and R, i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates were much larger with observed dissolved oxygen. Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates reduces tidal noise and produces more stable estimates with smaller ranges, regardless of the method used.

@@

The EBASE R package can also accommodate missing observations in the input data. This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.).  The R package accommodates data gaps using linear interpolation for any required parameter used by EBASE.  This allows the Bayesian routine to provide continuous estimates across the time series.  The results are then post-processed such that any metabolic estimates that exceed a threshold for interpolated observations are removed from the output.  The analyst can determine the appropriate length of time to use for excluding results based on interpolated values.  The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours.  Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis.  This feature allows EBASE to be used more readily because an analyst does not need to manually screen data that include missing values.

@@

```{r}
#| label: tbl-apacmptab
#| tbl-cap: "Summary statistics of daily metabolic estimates comparing the EBASE and Odum methods for net ecosystem metabolism (NEM), gross primary production (P), respiration (R), and gas exchange (D). Separate comparisons are made for estimates based on observed and detided dissolved oxygen. Summary statistics include Pearson correlation coefficients and root mean square error (RMSE) of a linear fit between the same metabolic estimates obtained from each method."
apacmptab
```
